---
title: "v3.1-stratifications Validation"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
author: Jenny McDaniel
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

<!-- # Purpose -->

<!-- This markdown will be used to validate stratification BED files for the -->
<!-- v3.1-stratification release. Validation includes: -->

<!-- 1. Perform sanity checks for each stratification file. -->
<!-- 2. Confirm expected chromosomes are covered and coverage of individual -->
<!--    chromosomes is as expected. -->
    
```{r, include=FALSE}
library(tidyverse)

coverage_plot <- function(df, level) {
  cat(sprintf("#### %s\n\n", level))
  p <- df %>%
    ggplot(aes(chrom, coverage)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(legend.position = "top",
          text = element_text(size = 30),
          axis.text.x = element_text(angle = 90),
          strip.text.y = element_text(angle = 0),
          panel.spacing.y = unit(2, "lines"))
  print(p)
  cat("\n\n")
}

coverage_plots <- function(df, type) {
  .df <- df %>% filter(strat_type == type)
  if (nrow(.df) == 0) {
    cat("None\n\n")
  } else {
    .df %>%
      group_by(strat_level) %>%
      group_walk(coverage_plot)
  }
}

knitr::opts_chunk$set(fig.width=25, fig.height=5, fig_caption=TRUE)

bed_cols <- c("chrom", "start", "end")

df_chr_sizes <- snakemake@input$nonN %>%
  read_tsv(col_names = bed_cols, col_types = "cii") %>%
  mutate(region_size = end - start) %>%
  group_by(chrom) %>%
  summarise(nonN_ref_len = sum(region_size))

read_strat <- function(path) {
  read_tsv(path,
           comment = "#",
           col_types = "cdd",
           col_names = bed_cols) %>%
    mutate(path = path)
}

df_coverage <- snakemake@input$strats %>%
  readLines() %>%
  as.list() %>%
  map_dfr(read_strat) %>%
  mutate(region_size = end - start) %>%
  group_by(path, chrom) %>%
  summarise(total_region = sum(region_size)) %>%
  # right join since we assume size df will have all desired chromosomes
  right_join(df_chr_sizes, by = "chrom") %>%
  mutate(strat_type = basename(dirname(path)),
         strat_filename = basename(path),
         genome = str_extract(strat_filename, "^[^_]+_(HG[0-9]+)_.bed.gz", 1),
         ## reference = str_extract(strat_filename, "^([^_])_.*.bed.gz", 1),
         strat_level = str_extract(strat_filename, 
                                   if_else(is.na(genome),
                                           "^[^_]+_(.*).bed.gz",
                                           "^[^_]+_HG[0-9]+_(.*).bed.gz"),
                                   1),
         coverage = total_region / nonN_ref_len) %>%
  select(-path)

included_chroms <- unique(df_coverage$chrom)
expected_chroms <- snakemake@params$expected_chroms

```

# Overview

```{r, echo = FALSE, results = "asis"}

cat(sprintf("Reference: %s\n\n", snakemake@wildcards$ref_key))

cat(sprintf("Build: %s\n\n", snakemake@wildcards$build_key))

cat(sprintf("Included chromosomes: %s\n\n", str_c(expected_chroms, collapse = ",")))

# TODO can also print out other build information here if we feel like it

```

# Coverage Plots

<!-- TODO these little panels can be grouped by "genome specific" and "everything" -->
<!-- else, and each can be generated in its own code block to keep the rmd DRY -->

## Low Complexity

Regions with different types and sizes of low complexity sequence (e.g.,
homopolymers, STRs, VNTRs, other locally repetitive sequences).

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "LowComplexity")
```

## Functional Technically Difficult

Different functional, or potentially functional, regions that are also likely to
be technically difficult to sequence.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "FunctionalTechnicallyDifficult")
```

## Functional Regions

Regions to stratify variants inside and outside of coding regions.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "FunctionalRegions")
```

## GC Content

Regions with different ranges(%) of GC content.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "GCcontent")
```

## Mappability

Regions where short read mappability can be challenging.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "mappability")
```

## Other Difficult

Highly variable regions like the VDJ and MHC, near gaps in the reference, or
errors in the reference.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "OtherDifficult")
```

## Segmental Duplications

Regions with segmental duplications or regions with non-trivial self-chain
alignments.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "SegmentalDuplications")
```

## Union

Regions with different general types of difficult regions or any type of
difficult region or complex variant. For example, performance can be measured in
just the "easy" regions of the genome.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "Union")
```

## Ancestry

Regions with inferred patterns of local ancestry in GRCh38.

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "ancestry")
```

## XY

Regions in chromosomes X and Y

```{r, echo = FALSE, results = "asis"}
coverage_plots(df_coverage, "XY")
```

## Genome Specific (benchmark v4.2.1)

Difficult regions due to potentially difficult variation in the NIST/GIAB
sample, including:

1. regions containing putative compound heterozygous variants
2. regions containing multiple variants within 50 bp of each other
3. regions with potential structural variation and copy number variation

```{r, echo = FALSE, results = "asis"}
# TODO this will need to be broken out by specific genomes in the plot
coverage_plots(df_coverage, "GenomeSpecific")
```
